---
title: "Trends and Determinants of Obesity in Scotland: An Analysis of the 2008-2012 Scottish Health Surveys"
format: html
eval: true
echo: false
editor: visual
---

```{r}
#| warning: false
#| message: false

library(ggplot2)
library(tidyverse)
library(tidymodels)
library(gt)
library(MASS)
library(patchwork)
library(moderndive)
library(dplyr)
library(gapminder)
library(sjPlot)
library(stats)
library(readr)
library(janitor)

```

```{r}
#|warning: false
#|message: false

obese <- read_csv("H:/DAProject10 (1).csv")

obese$Sex <- as.factor(obese$Sex)

obese<- obese|>mutate(obese_indicator = ifelse(BMIgroup == "Obese", 1, 0))

obese$obese_indicator <- as.factor(obese$obese_indicator)

```

# Formal Analysis {#sec-FA}

Now that we have carried out exploratory analysis and we can see the trends in the data we want to fit a logistic regression model to the data.

First we will analyse the obesity trends over the years.

```{r}
model.years <- logistic_reg() |> 
  set_engine("glm")
model.years <- model.years |> 
  fit(obese_indicator ~ Year, data = obese) |>
  extract_fit_engine()

model.years |>
  summary()
```


Here we have fitted a logistic regression model with obesity as the response variable and year as the explanatory variable.

The baseline category for our binary response is 0 which corresponds to the subject not being obese. This means that the estimates from the logistic regression model are for a change in the log odds scale for being obese in comparison to the response baseline which is not being obese.


$$ ln\frac{p}{1-p}=\alpha +\beta \cdot year = -12.1 + 0.005 \cdot year $$ 

where:

- $p=Prob(Not obese)$ and
- $1-p = Prob(Obese)$

Hence, the log odds of the person not being obese increased by 0.005 for every one unit increase in year. However, the coefficient of year has a p value of 0.586 which means that it is not significant.


```{r}

plot_model(model.years, show.values = TRUE,
           title = "Odds (year)", show.p = TRUE)
```

This plot shows the odds of being obese as the years go on.

However, if we still want to use this coefficient to quantify the effect of time we would need to work out the odds scale as it is easier to interpret. to find this we simply exponentiate the log-odds:

$$\frac{p}{1-p}= exp(\alpha+\beta \cdot age)$$
So if we take exp(0.005)=1.05 This suggests the odds of a person not being obese is 1.05 times greater in a later year compared to an earlier one.



```{r}
plot_model(model.years, type = "pred", terms = "Year", title = "Probability of obesity over time", axis.title = c("year", "Probability of being obese"))
```

from this plot we can see the probability someone is not obese as time goes on.

Now that we have analysed the relationship between obesity over time we can carry out formal analysis for the relationship between obesity as our response variable and Age, Sex and education as our explanatory variables.

Again we will fit a logistic regression model.


