---
title: "Trends and Determinants of Obesity in Scotland: An Analysis of the 2008-2012 Scottish Health Surveys"
format: html
eval: true
echo: false
editor: visual
number-sections: true
---

```{r}
#| warning: false
#| message: false

library(ggplot2)
library(tidyverse)
library(tidymodels)
library(gt)
library(MASS)
library(patchwork)
library(moderndive)
library(dplyr)
library(gapminder)
library(sjPlot)
library(stats)
library(readr)
library(janitor)

```

```{r}
#|warning: false
#|message: false

obese <- read_csv("H:/DAProject10 (1).csv")

obese$Sex <- as.factor(obese$Sex)

obese<- obese|>mutate(obese_indicator = ifelse(BMIgroup == "Obese", 1, 0))

obese$obese_indicator <- as.factor(obese$obese_indicator)

```

# Formal Analysis {#sec-FA}

Now that we have carried out exploratory analysis and we can see the trends in the data we want to fit a logistic regression model to the data.

## 3.1 Has the prevalence of obesity in Scotland changed over the given years?

First we will analyse the obesity trends over the years.

```{r}
#| eval: false
model.years <- logistic_reg() |> 
  set_engine("glm")
model.years <- model.years |> 
  fit(obese_indicator ~ Year, data = obese) |>
  extract_fit_engine()

model.years |>
  summary()
```


|               | Estimate   | P value |
|---------------|------------|---------|
| **Intercept** | -12.130043 | 0.558   |
| **Year**      | 0.005604   | 0.586   |

Here we have fitted a logistic regression model with obesity as the response variable and year as the explanatory variable.

The baseline category for our binary response (intercept) is 0 which corresponds to the subject not being obese. This means that the estimates from the logistic regression model are for a change in the log odds scale for being obese in comparison to the response baseline which is not being obese.

Hence, the log odds of the person not being obese increased by 0.005 for every one unit increase in year. However, the coefficient of year has a p value of 0.586 which means that it is not significant. Hence, there is not a significant change in obesity levels over the years given in the study.

$$ ln\frac{p}{1-p}=\alpha +\beta \cdot year = -12.1 + 0.005 \cdot year $$ 

where:

- $p=Prob(Not obese)$ and
- $1-p = Prob(Obese)$








```{r}
plot_model(model.years, type = "pred", terms = "Year", title = "Probability of obesity over time", axis.title = c("year", "Probability of being obese"))
```

from this plot we can see the probability someone is not obese as time goes on.

## 3.2 Are there any differences in obesity by age, gender and education?


Now that we have seen there is no relationship between obesity levels and the given time frame of our study we can carry out formal analysis for the relationship between obesity as our response variable and Age, Sex and Education as our explanatory variables.

Again we will fit a logistic regression model, this time with obesity as our response variable and age, sex and highest education qualification as our explanatory variables.
```{r}
#|eval: false
mod2 <- logistic_reg()
mod2 <- mod2 |> fit (obese_indicator~ Sex + Age + Education, data = obese) |>
  extract_fit_engine()
mod2 %>%
  summary()

```

|                                         | Estimate  | P Value    |
|-----------------------------------------|-----------|------------|
| Intercept                               | -1.647009 | \<2e\^-16  |
| SexMale                                 | -0.029207 | 0.29681    |
| Age                                     | 0.011276  | \<2e\^-16  |
| Education- Higher Grade or Equivalent   | 0.158802  | 0.00068    |
| Education- HNC/D or equivalent          | 0.249311  | 1.52e\^0.6 |
| Education- Standard grade or equivalent | 0.323521  | 2.02e\^-14 |
| Education- No qualifications            | 0.409377  | \<2e\^-16  |
| Education- other                        | 0.234054  | 6.17e\^0.5 |

We can see that the coefficient for males is negative indicating a lower chance of males not being obese. However, the p value is 0.29681 which indicates that this variable is insignificant and obesity levels do not differ by gender.

The age coefficient is positive suggesting that older people were more likely to be obese. This has a low p value so is significant in our analysis.

We can also see that the coefficients for all of the levels of education are positive, with individuals who have no qualifications having the highest coefficient. This suggests that individuals with no qualifications have a highest chance of being obese compared to those who have a degree qualification. We can see that those with a Standard grade qualification have the second highest chance of being obese compared to those with a degree, followed by those with an other qualification and those with a standard grade qualification. The group least likely to be obese compared to those with a degree are the people that have a higher qualification or equivalent.



```{r}
#| fig-cap: Odds of being obese 
#| label:  fig-odds
plot_model(mod2, show.values = TRUE,
           title = "", show.p = FALSE, value.offset = 0.25)
```

We can interpret these odds ratios. We know sex is an insignificant factor.

For each year increase in an individuals age, their odds of being obese increase by 1.01.

The plot suggests that People with a higher qualification or equivalent's odds of being obese are 1.17 times higher than those who have a degree.

The odds of people with a HNC/D being obese is 1.28 times that of those who have a degree.

The odds of people with No qualifications being obese is 1.51 times those who have a degree.

The odds of an individual who has another level of qualification being obese is 1.26 times those who have a degree, an finally the odds of those who have a standard grade or equivalent qualification being obese is 1.38 times those who have a degree.





